---
title: "NYPDShooting"
author: "C. Knowles"
date: "2023-11-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo=FALSE, include=FALSE}
library(hms)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
```


## NYPD Shooting l  
### Cleaning the Data

```{r read and show summary, echo=TRUE} 
shooting_csv <- read.csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
head(shooting_csv)
summary(shooting_csv)
str(shooting_csv)
```
Here my approach is to first remove any columns that will be unnecessary for analysis - LOC_OF_OCCUR_DESC, PRECINCT, JURISDICTION_CODE,  LOC_CLASSFCTN_DESC, LOCATION_DESC, STATISTICAL_MURDER_FLAG, X_COORD_CD, Y_COORD_CD, Latitude, Longitude, and Lon_Lat.  

```{r remove columns, echo=TRUE}
shoot <- distinct(shooting_csv) %>% 
          select(OCCUR_DATE, OCCUR_TIME, BORO, PERP_AGE_GROUP, PERP_SEX, PERP_RACE, VIC_AGE_GROUP, VIC_SEX, VIC_RACE)
```  

Next, I want to remove any rows that contain typos, missing, or 'UNKNOWN' data. I chose not to impute missing data because all missing values are strictly categorical and trying to predict what is missing is futile. 


```{r remove rows with missing or UNKNOWN data, echo=TRUE}
filter_values <- c("", "UNKNOWN", "(null)", "U")
cleaned_shoot <- shoot %>% filter(!(BORO %in% filter_values) & 
                                    !(PERP_AGE_GROUP %in% c(filter_values, "1020", "224", "940")) & 
                                    !(PERP_SEX %in% filter_values) & 
                                    !(PERP_RACE %in% filter_values) & 
                                    !(VIC_AGE_GROUP %in% c(filter_values, "1022")) & 
                                    !(VIC_SEX %in% filter_values) & 
                                    !(VIC_RACE %in% filter_values))

#cleaned_shoot <- cleaned_shoot %>% filter(PERP_AGE_GROUP == 1 & VIC_AGE_GROUP == 1)
```  

Now that my data set is cleaned, I want to transform it into a format that will be easier to work with going forward. I started by converting the OCCUR_DATE column to a date type, the OCCUR_TIME column to a time type, and the remaining columns to a factor type. This will allow me to easily sort and group our categorical data for visuals.

```{r convert to factor/date/time, echo=TRUE}
glimpse(cleaned_shoot)
cols_to_convert <- c("BORO", "PERP_AGE_GROUP", "PERP_SEX", "PERP_RACE", "VIC_AGE_GROUP", "VIC_SEX", "VIC_RACE")
cleaned_shoot[cols_to_convert] <- lapply(cleaned_shoot[cols_to_convert], as.factor)
cleaned_shoot$OCCUR_DATE <- as.Date(cleaned_shoot$OCCUR_DATE, format = "%m/%d/%Y")
cleaned_shoot$OCCUR_TIME <- as_hms(cleaned_shoot$OCCUR_TIME)
shoot_tibble <- as_tibble(cleaned_shoot)
shoot_tibble
```  

The following questions will be explored in this analysis  

1) Does one race make up the majority of perpetrators?  
  1b) ... majority of victims?  
  1c) Is this proportional to a borough's racial demographic?  
2) Do shootings occur significantly more among a specific age group?  
  2b) ... among sex?  
  2c) ... among race?  
3) Do shootings occur significantly more in one month/s versus others?  
  3b) ... in one time of day versus others?  
4) Are there any confounding variables we must consider?  

```{r frequencies}
boro_counts <- count(shoot_tibble, BORO)
print(boro_counts)

ggplot(shoot_tibble, aes(x = BORO)) +
  geom_bar() +
  labs(title = "Distribution of Incidents by Borough")
cross_tab <- table(shoot_tibble$BORO, shoot_tibble$PERP_AGE_GROUP)
print(cross_tab)

ggplot(shoot_tibble, aes(x = PERP_AGE_GROUP)) +
  geom_bar() + 
  labs(title = "Count of reported incidents grouped by perpetrator age")

ggplot(shoot_tibble, aes(x = VIC_AGE_GROUP)) +
  geom_bar() + 
  labs(title = "Count of reported incidents grouped by victim age")

chi_sqaure_result <- chisq.test(shoot_tibble$BORO, shoot_tibble$PERP_AGE_GROUP)
print(chi_sqaure_result)
contingency_table <- table(shoot_tibble$BORO, shoot_tibble$PERP_AGE_GROUP)

fisher_result <- fisher.test(contingency_table, simulate.p.value = TRUE)
print(fisher_result)

chi_sqaure_result2 <- chisq.test(shoot_tibble$BORO, shoot_tibble$VIC_AGE_GROUP)
print(chi_sqaure_result2)

chi_sqaure_result3 <- chisq.test(shoot_tibble$PERP_AGE_GROUP, shoot_tibble$VIC_AGE_GROUP)
print(chi_sqaure_result3)
```
For each chi square test we reject the null hypothesis, therefore, there is a relationship between BORO v.s. PERP_AGE_GROUP, BORO v.s. VIC_AGE_GROUP, and PERP_AGE_GROUP v.s. VIC_AGE_GROUP.


